### 2018-03-27 submission/manuscript.Rmd
add better transitions to intro

A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature has been shown to cause overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination has been shown to add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Recent studies have shown that the majority of these biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011].

There are also intrinsic properties to primers and DNA polymerases chosen that can introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

@@

Although PCR has been investigated as a source of bias in 16S rRNA gene sequencing, the contribution of specific DNA polymerases has not been. This study will specifically address how high fidelity (HiFi) DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

>>>>

Many parts of the 16S rRNA gene sequencing workflow contribute bias to the results and are studied extensively. A typical 16S rRNA gene sequencing workflow can be divided into preservation, extraction, PCR, and sequencing steps. Generally, not using a preservation media and leaving samples at room temperature supports overgrowth of low abundance members of the fecal bacterial community [@Amir2017]. Similarly, this overgrowth can still occur if the preservation media does not adequately inhibit growth [@preservation_Luo_2016; @preservation_Song_2016]. Reports have also shown that changes in specific community members might occur due to differing susceptibility to freeze thaw cycles amongst microbes [@Gorzelak2015]. Additionally, reagent contamination can add community members and the contribution of these contaminant members grows larger with lower biomass samples [@contamination_Salter2014]. Overall, biases due to either preservation or extraction tend to be smaller than the overall biological signal being measured [@preservation_Song_2016; @storage_Bassis_2017]. However, the contribution of PCR bias to this overall workflow is not well characterized since these studies use the same PCR approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is important because a large body of literature shows that there are a variety of steps during PCR that can change the observed results [@Eckert1991; @Burkardt2000]. Many of these sources of biases are made worse as cycle number increases [@Wang1996; @Haas2011; @Kebschull2015]. For example, the selective amplification of AT-rich over GC-rich sequences can exaggerate the difference between 16S rRNA genes higher in AT versus those higher in GC [@polz_bias_1998]. Both amplification error and non-specific amplification (e.g. incorrect amplicon size products) can also increase as cycle number increases which can drastically change commonly used diversity measures [@BautistadelosSantos2016; @Acinas2005]. Additionally, chimeras can form from an aborted extension step followed by a subsequent priming error and secondary extension and will also artificially increase community diversity [@Haas2011]. Although these differences are not necessarily dependent on primer and DNA polymerase used, there are also biases that are.

The intrinsic properties to primers and DNA polymerases chosen can also introduce bias. Primers have variable region dependent binding affinities for different bacteria and depending on the primer pair do not detect specific bacteria (e.g. V1-V3 does not detect *Haemophilus influenzae* and V3-V5 does not detect *Propionibacterium acnes*) [@Meisel2016; @Sze2015 (Table S4)]. Additionally, there are multiple families of DNA polymerases that have their own error rate and proof reading capacity [@polymerase_Ishino_2014]. Interestingly, the influence that these different DNA polymerases can have on the observed 16S rRNA gene sequencing results have not been well studied like some of the other previously mentioned sources of PCR-based bias.

@@

Although bias introduced due to differences between DNA polymerase and HiFi DNA polymerase has been investigated for 16S rRNA gene sequencing, the bias caused due to differences between specific HiFi DNA polymerases has not been. This study will specifically address how HiFi DNA polymerases can bias observed bacterial community results derived from 16S rRNA gene sequencing. We will accomplish this by examining if any of five different types of HiFi DNA polymerases introduce significant biases into 16S rRNA gene surveys, if this is a cycle dependent phenomenon, and whether they can be removed using a standard bioinformatic pipeline.

### 2018-03-27 submission/manuscript.tex
add better transitions to intro

Accounting for bias is critical in order to reach an understanding of
bacterial community changes using 16S rRNA gene sequencing results.
First, it is important to differentiate between bias, reproducibility,
and standardization since often times these three can be confused and
used interchangeably with each other. Bias can change the observed
results in a way that is reproducible and standardized. For example, if
one group uses one brand of DNA extraction kit for their 16S rRNA gene
sequencing, their results may be biased versus another group not using
the same brand kit but within their group they can still have
reproducible results. Therefore, standardization of 16S rRNA gene
sequencing methods to increase reproducibility can still be problematic
due to bias. Determining how different aspects of a 16S rRNA gene
sequencing workflow could bias the observed results is critical for the
interpretation of specific studies in the broader context of the overall
field.

A typical 16S rRNA gene sequencing workflow can be divided into
preservation, extraction, PCR, and sequencing steps. Generally, not
using a preservation media and leaving samples at room temperature has
been shown to cause overgrowth of low abundance members of the fecal
bacterial community (Amir et al., 2017). Similarly, this overgrowth can
still occur if the preservation media does not adequately inhibit growth
(Song et al., 2016; Luo et al., 2016). Reports have also shown that
changes in specific community members might occur due to differing
susceptibility to freeze thaw cycles amongst microbes (Gorzelak et al.,
2015). Additionally, reagent contamination has been shown to add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Recent
studies have shown that the majority of these biases due to either
preservation or extraction tend to be smaller than the overall
biological signal being measured (Song et al., 2016; Bassis et al.,
2017). However, the contribution of PCR bias to this overall workflow is
not well characterized since these studies use the same PCR approach
while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will also
artificially increase community diversity (Haas et al., 2011).

There are also intrinsic properties to primers and DNA polymerases
chosen that can introduce bias. Primers have variable region dependent
binding affinities for different bacteria and depending on the primer
pair do not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other previously mentioned
sources of PCR-based bias.

>>>>

Accounting for bias is critical in order to reach an understanding of
bacterial community changes using 16S rRNA gene sequencing results.
Differentiating between bias, reproducibility, and standardization is
important since often times these three can be confused and used
interchangeably with each other. Bias can change the observed results in
a way that is reproducible and standardized. For example, if one group
uses one brand of DNA extraction kit for their 16S rRNA gene sequencing,
their results may be biased versus another group not using the same
brand kit but within their group they can still have reproducible
results (Salter et al., 2014). Our 16S rRNA gene sequencing methods are
biased even when these workflows are standardized to increase
reproducibility. In order to interpret specific studies within the
broader context of the overall field, assessing bias at different parts
of the 16S rRNA gene sequencing workflow is critical.

Many parts of the 16S rRNA gene sequencing workflow contribute bias to
the results and are studied extensively. A typical 16S rRNA gene
sequencing workflow can be divided into preservation, extraction, PCR,
and sequencing steps. Generally, not using a preservation media and
leaving samples at room temperature supports overgrowth of low abundance
members of the fecal bacterial community (Amir et al., 2017). Similarly,
this overgrowth can still occur if the preservation media does not
adequately inhibit growth (Song et al., 2016; Luo et al., 2016). Reports
have also shown that changes in specific community members might occur
due to differing susceptibility to freeze thaw cycles amongst microbes
(Gorzelak et al., 2015). Additionally, reagent contamination can add
community members and the contribution of these contaminant members
grows larger with lower biomass samples (Salter et al., 2014). Overall,
biases due to either preservation or extraction tend to be smaller than
the overall biological signal being measured (Song et al., 2016; Bassis
et al., 2017). However, the contribution of PCR bias to this overall
workflow is not well characterized since these studies use the same PCR
approach while varying preservation or extraction method.

Identifying the biases in the PCR stage of 16S rRNA gene sequencing is
important because a large body of literature shows that there are a
variety of steps during PCR that can change the observed results (Eckert
\& Kunkel, 1991; Burkardt, 2000). Many of these sources of biases are
made worse as cycle number increases (Wang \& Wang, 1996; Haas et al.,
2011; Kebschull \& Zador, 2015). For example, the selective
amplification of AT-rich over GC-rich sequences can exaggerate the
difference between 16S rRNA genes higher in AT versus those higher in GC
(Polz \& Cavanaugh, 1998). Both amplification error and non-specific
amplification (e.g.~incorrect amplicon size products) can also increase
as cycle number increases which can drastically change commonly used
diversity measures (Acinas et al., 2005; Santos et al., 2016).
Additionally, chimeras can form from an aborted extension step followed
by a subsequent priming error and secondary extension and will also
artificially increase community diversity (Haas et al., 2011). Although
these differences are not necessarily dependent on primer and DNA
polymerase used, there are also biases that are.

The intrinsic properties to primers and DNA polymerases chosen can also
introduce bias. Primers have variable region dependent binding
affinities for different bacteria and depending on the primer pair do
not detect specific bacteria (e.g.~V1-V3 does not detect
\emph{Haemophilus influenzae} and V3-V5 does not detect
\emph{Propionibacterium acnes}) (Sze et al., 2015 (Table S4); Meisel et
al., 2016). Additionally, there are multiple families of DNA polymerases
that have their own error rate and proof reading capacity (Ishino \&
Ishino, 2014). Interestingly, the influence that these different DNA
polymerases can have on the observed 16S rRNA gene sequencing results
have not been well studied like some of the other previously mentioned
sources of PCR-based bias.
